---
title: "Migration Phenology"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(lubridate)
require(tidyr)
require(vegan)
require(ggplot2)
```

**II. Decomposing the Diversity Index Trends**  

**A. Goshute Mts (West)**  

```{r include=F}
GosMts=read.csv("GosMts.csv")

gos=GosMts%>%select(-TOTAL)%>%
  pivot_longer(cols=3:27, names_to = "Species", values_to="Count")%>%
  mutate(year=year(Date), month=month(Date), date=day(Date))%>%
  filter(!is.na(Count))

gos=gos%>%mutate(Julian=as.POSIXlt(Date)$yday,
                  diet=case_when(Species=="AK" ~"bird",
                                 Species=="BE" ~"other",
                                 Species=="CH" ~"bird",
                                 Species=="BV" ~ "other",
                                 Species=="TV" ~"other",
                                 Species=="OS" ~"other",
                                 Species== "NH" ~"mammal",
                                 Species=="SS" ~"bird",
                                 Species=="NG"~ "bird",
                                 Species=="RS" ~"mammal",
                                 Species=="BW"~"mammal",
                                 Species=="RT" ~"mammal",
                                 Species=="RL" ~"mammal",
                                 Species=="SW"~"mammal",
                                 Species=="FH"~"mammal",
                                 Species=="GE"~"mammal",
                                 Species=="ML"~"bird",
                                 Species=="PG"~"bird",
                                 Species=="PR"~"mammal",
                                 Species=="MK"~"other"))%>%
                   filter(!Species%in%c("UA", "UB", "UF", "UE", "UR"))

gos_sp_max=gos%>%mutate(Julian=as.POSIXlt(Date)$yday)%>%
    group_by(year, Species, diet)%>%
    filter(Count==max(Count), !Species%in%c("UA", "UB", "UE", "UF", "UR"))%>%slice_head(n=1)%>%
  arrange(year)

gos_sp=gos%>%mutate(Julian=as.POSIXlt(Date)$yday)%>%
  group_by(year, Species, diet)%>%
  summarise(yr_tot=sum(Count))%>%filter(!Species%in%c("UA", "UB", "UE", "UF", "UR"))%>%
  arrange(year)

#DAILY TOTALS
gostot=GosMts%>%select(-TOTAL)%>%
  mutate(yr_tot=rowSums(.[3:27], na.rm=T), year=year(Date), month=month(Date), date=day(Date))%>%
  filter(!year>2018)
gostot$Julian=as.POSIXlt(gostot$Date)$yday 

gostotmax=gostot%>%group_by(year)%>%filter(yr_tot==max(yr_tot))
```  

```{r echo=F, message=F}
#site
ggplot(gostotmax, aes(x=year, y=Julian))+geom_point()+geom_hline(yintercept=268, linetype=2)+ggtitle("Goshutes")+ylab("peak migration date(Julian)")+
  stat_smooth(method="lm")+theme_classic()

#species
ggplot(gos_sp_max, aes(x=year, y=Julian, col=Species))+geom_point()+facet_wrap(~diet)+
  stat_smooth(method="lm")+geom_hline(yintercept=268, linetype=2)+ylab("peak migration date(Julian)")+
  theme_classic()

ggplot(gos_sp, aes(x=year, y=log(yr_tot), col=Species))+geom_point()+facet_wrap(~diet)+
  stat_smooth(method="lm")+theme_classic()+ylab ("total counts (log)")
```  

**B. Hawk Mt. (east)**  

```{r echo=F, include=F}
hms=read.csv("HMSDay.csv")
#combine baea and goea data
hmstot=hms%>%
  mutate(BAEA=rowSums(.[9:11]), GOEA=rowSums(.[21:23]))%>%
  select(-OBSERVERS, -GOEA.I, -GOEA.A, -GOEA.U, -BAEA.I, -BAEA.U, -BAEA.A, -RAPTORS)%>%
  mutate(yr_tot=rowSums(.[6:30], na.rm=T), date=make_date(YR, MO, DAY))%>%filter(!YR<1983, !YR>2018)

hmstot$Julian=as.POSIXlt(hmstot$date)$yday 

#species-level####

hms_sp=hmstot%>%select(-yr_tot)%>%
  pivot_longer(cols=6:30, names_to = "Species", values_to="Count")%>%
  filter(!is.na(Count), !YR<1983, !Species%in%c("UNID.ACCIPITER",
                                                "UNID.BUTEO", 
                                                "UNID.EAGLE",
                                                "UNID.FALCON",
                                                "UNID.RAPTOR"))%>%
  mutate(Date=make_date(YR, MO, DAY))

hms_sp=hms_sp%>%mutate(diet=case_when(Species=="AMKE" ~"bird",
                                 Species=="BAEA" ~"other",
                                 Species=="COHA" ~"bird",
                                 Species=="BLVU" ~ "other",
                                 Species=="TUVU" ~"other",
                                 Species=="OSPR" ~"other",
                                 Species== "NOHA" ~"mammal",
                                 Species=="SSHA" ~"bird",
                                 Species=="NOGO"~ "bird",
                                 Species=="RSHA" ~"mammal",
                                 Species=="BWHA"~"mammal",
                                 Species=="RTHA" ~"mammal",
                                 Species=="RLHA" ~"mammal",
                                 Species=="GYRF"~"mammal",
                                 Species=="SWHA"~"mammal",
                                 Species=="FEHA"~"mammal",
                                 Species=="GOEA"~"mammal",
                                 Species=="MERL"~"bird",
                                 Species=="PEFA"~"bird",
                                 Species=="PRFA"~"mammal",
                                 Species=="MIKI"~"other",
                                 Species=="STKI" ~"other"))
#PEAK DAYS PER SPECIES
hms_sp_max=hms_sp%>%mutate(Julian=as.POSIXlt(Date)$yday)%>%
  group_by(YR, Species,diet)%>%
  filter(Count==max(Count))%>%slice_head(n=1)%>%
  arrange(YR)

hm_spp=hms_sp%>%group_by(Species,YR,diet)%>%summarise(yr_tot=sum(Count))

#PEAK DAYS FOR SITE
hmsmax=hmstot%>%
  select(-UNID.ACCIPITER, -UNID.BUTEO, -UNID.EAGLE, -UNID.FALCON, -UNID.RAPTOR)%>%
  mutate(Julian=as.POSIXlt(date)$yday)%>%
  group_by(YR)%>%
  filter(yr_tot==max(yr_tot))%>%
  arrange(YR)
```

```{r echo=F, warning=F, message=F}

#site
ggplot(hmsmax, aes(x=YR, y=Julian))+geom_point()+stat_smooth(method="lm")+theme_classic()+
  geom_hline(yintercept=260, linetype=2)+ylab("peak migration date(Julian)")

ggplot(hms_sp_max, aes(x=YR, y=Julian, col=Species))+geom_point()+stat_smooth(method="lm")+theme_classic()+
  facet_wrap(~diet)+geom_hline(yintercept=260, linetype=2)+ylab("peak migration date(Julian)")

ggplot(hm_spp, aes(x=YR, y=log(yr_tot),col=Species))+geom_point()+stat_smooth(method="lm")+theme_classic()+
  facet_wrap(~diet)+ylab("total counts (log)")
```  

**Hawk Ridge (central)**

```{r include=F}
HawRd=read.csv("HawkRid.csv")
HawRd[, 3:24][is.na(HawRd[, 3:24])] <- 0

#without unknowns
hrtot=HawRd%>%select(-TOTAL)%>%
  mutate(yr_tot=rowSums(.[3:19], na.rm=T), year=year(Date), month=month(Date), date=day(Date))%>%
  filter(!year>2018) #get annual totals across all spp.

hrtot$Julian=as.POSIXlt(hrtot$Date)$yday 

#species-level####

hr_sp=hrtot%>%
  pivot_longer(cols=3:24, names_to = "Species", values_to="Count")%>%
  filter(!is.na(Count))

hr_sp=hr_sp%>%mutate(
                  diet=case_when(Species=="AK" ~"bird",
                                 Species=="BE" ~"other",
                                 Species=="CH" ~"bird",
                                 Species=="BV" ~ "other",
                                 Species=="TV" ~"other",
                                 Species=="OS" ~"other",
                                 Species== "NH" ~"mammal",
                                 Species=="SS" ~"bird",
                                 Species=="NG"~ "bird",
                                 Species=="RS" ~"mammal",
                                 Species=="BW"~"mammal",
                                 Species=="RT" ~"mammal",
                                 Species=="RL" ~"mammal",
                                 Species=="SW"~"mammal",
                                 Species=="FH"~"mammal",
                                 Species=="GE"~"mammal",
                                 Species=="ML"~"bird",
                                 Species=="PG"~"bird",
                                 Species=="PR"~"mammal",
                                 Species=="MK"~"other"))%>%
                   filter(!Species%in%c("UA", "UB", "UF", "UE", "UR"))


hr_sp_max=hr_sp%>%group_by(Species,year,diet)%>%filter(Count==max(Count))%>%slice_head(n=1)%>%select(-yr_tot)

hr_spp=hr_sp%>%group_by(Species,year,diet)%>%summarise(yr_tot=sum(Count))

hrmax=hrtot%>%group_by(year)%>%filter(yr_tot==max(yr_tot))%>%slice_head(n=1)%>%select(-UA, -UB,-UE,-UF,-UR)%>%
  mutate(Julian=as.POSIXlt(Date)$yday )
```  

```{r echo=F, warning=F, message=F}
ggplot(hrmax, aes(x=year, y=Julian))+geom_point()+stat_smooth(method="lm")+theme_classic()+
  geom_hline(yintercept=257, linetype=2)+ylab("peak migration date(Julian)")

ggplot(hr_sp_max, aes(x=year, y=Julian, col=Species))+geom_point()+stat_smooth(method="lm")+theme_classic()+
  facet_wrap(~diet)+geom_hline(yintercept=257, linetype=2)+ylab("peak migration date(Julian)")

ggplot(hr_spp, aes(x=year, y=log(yr_tot), col=Species))+geom_point()+stat_smooth(method="lm")+theme_classic()+
  facet_wrap(~diet)+ylab("total counts (log)")
```   

**Questions for Morgan**  

1. would it be a reasonable approach, instead of binning species into these functional groups, I use continuous values of % of their diet being insects,birds, mammals,etc? I'm thinking of looking at whether or not species that have higher %s of their diets being mammals/birds are advancing/delaying their timing or increasing/decreasing in abundances?  that way I don't have to bin together ospreys and vultures.Although, if I do want to look at how functional groups are driving temporal diversity patterns at the site level, I would probably want to have a binned value of the timing and abundances of these groups,right?

2. would regression models for the two main values at the "assemblage"/(aka site)-level (i.e., H' index values and peak migration dates) and at the species-level (i.e., annual abundances and peak migration dates) be appropriate to test the trends?  

3. what other exploratory work should I consider doing?  

